---
title: "salmon_permit_theory"
author: "Ben Buzzee"
date: "February 12, 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

All data is taken from the permit_records and harvest_records sheets in the original excel document. First we will take note of the form of the data we have:

```{r}
# Script to recreate the 2017 Salmon Harvest estimates


#=====================================================================
# Read in and examine the data
#=====================================================================


library(tidyverse)

# Data Source: S:\RTS\common\Pat\Permits\Shrimp\2017\SHRIMP PERMITS 2017.XLS
# I saved individual copies of the two following sheets to my project folder

harvest <- read.csv("../data/harvest_records_17.csv", header = T, nrow = 32187)
permits <- read.csv("../data/permit_records_17.csv", header = T, nrow = 29584)

# change all variable names to lower (personal preference)
names(harvest) <- tolower(names(harvest))
names(permits) <- tolower(names(permits))

# make it so each row is one permit-fishery and their harvest totals
harvest <- harvest %>% group_by(permit, fishery) %>% summarise(red = sum(red), king = sum(king), coho = sum(coho), pink = sum(pink), chum = sum(chum))




# calculate counts for each mailing type as well as each mailing-status combination
# and verify that they match the counts in the excel document

mail_cnts <- permits %>% group_by(mailing) %>% summarise(count = n())
status_cnts <- permits %>% group_by(mailing, status) %>% summarise(count = n())

#sum across all dates to get total fishing effort per permit and fishery

mail_cnts <- rbind(mail_cnts, c("total", sum(mail_cnts$count)))

```



We have the total number of permits, and the mailing status of each:
```{r}
knitr::kable(mail_cnts)
```


And of all those that reported fishing, (mailing status 0, 1, or 2), we have the number of fish they caught at each location they fished at:

```{r}
knitr::kable(head(harvest))
```


Our goal is to estimate how many fish were harvested from each location by the 7268 permit holders that never responded.


First, we observe that the total number of permits issued can be split into three groups:


$N = N_c + N_2 + N_{nr}$


where:

* $N_c$ is the number of compliant households that responded without a reminder or after the first reminder (mailing = 0 or 1)
* $N_2$ is the number of households that responded after the second reminder (mailing = 2)
* $N_{nr}$ is the number of households that never responded. (mailing = 9)



We will assume that $N_2 + N_{nr}$ is a population of "noncompliant" households and $N_2$ is a random sample from said population. Further, note that $N_c$, $N_2$ and the harvests from those households are fixed and known. We only need to estimate the harvest of households that never responded.

Since the households making up $N_2$ are a random sample, both $N_2$ and $N_{nr}$ share the same population mean harvest and proportion of permit holders that fished. So we will use observations from $N_2$ to estimate those parameters. From the households making up $N_2$, we will use the estimates:

* $\hat{w}$ as the proportion of $N_2$ that participated in fishing: $\hat{w}$ = $\frac{n_{2f}}{N_2}$ where $n_{2f}$ is the number of households responding to the second reminder (mailing = 2) that reported fishing.
* and $\bar{h}_{l}$ as the average harvest at location $l$ __among all of those that fished__ : $\bar{h}_{l} = \frac{\sum_{i=1}^{n_{2f}} (h_{li})}{n_{2f}}$ over households 
$i=1, 2,..., n_{2f}$ and fixed location $l$

By the central limit theorem,

$\hat{w}$ follows a $N(w, \frac{\sigma^2}{N_2})$ distribution with $\hat{var}(\hat{w}) = \frac{\hat{w}(1-\hat{w})}{N_2}$

and $\bar{h_{l}}$ is $N(h_{l}, \frac{\sigma^2}{n_{2f}})$ with $\hat{var}(\bar{h_{l}}) = \frac{s^2_{l}}{n_{2f}}$ where $s^2_{l} = \Sigma_{i=1}^{n_{2f}}\frac{(h_{li} - \bar{h_{l}})^2}{n_{2f}-1}$



So the estimated number of non-respondents that fished is: $\hat{N}_{nrf} = N_{nr}\hat{w}$ and the estimated total harvest at each location by non-respondents is $\hat{H}_{nrfl} = \hat{N}_{nrf}\bar{h_{l}}$. Our estimated variance of the harvest estimate is then:

$\hat{var}(\hat{H}_{nrfl}) = \hat{var}(\hat{N}_{nrf}\bar{h_{l}}) =  \hat{var}({N_{nr}\hat{w}\bar{h_{l}})} = N_{nr}^2\hat{var}(\hat{w}\bar{h_{l}}) =$

$N_{nr}^2[\bar{h_{l}}^2\hat{var}(\hat{w}) + \hat{w}^2\hat{var}(\bar{h_{l}}) - \hat{var}(\hat{w})\hat{var}(\bar{h_{l}})]$ by Goodman (1960)

Finally, our estimated total harvest for each location would be $H_{cl} + H_{2l} + \hat{H}_{nrfl}$ where both $H_{cl}$ and $H_{2l}$ are known. Since they are both known, the total variance would be the same as above.