---
title: "Shrimp Permits"
author: "Ben Buzzee, Biometrician, ADFG"
date: "July 27, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# NOTE TO SELF: double  check that reported fishign == "harvest reported"
```



```{r init, echo = FALSE, include = FALSE}
library(tidyverse)

harvest <- read.csv("harvest_records_17.csv", header = T, nrows = 10687)
permits <- read.csv("permit_records_17.csv", header = T)

names(harvest) <- tolower(names(harvest))
names(permits) <- tolower(names(permits))
harvest$harvdate <- as.character(harvest$harvdate)

harvest <- harvest %>% mutate(reported_fishing = 1)


# rm_jay <- c(10733,
#             12311,
#             10733,
#             13037,
#             10414)
# 
# rm_date <- c("4/15/2017",
#              "7/5/2017",
#              "4/16/2017",
#              "7/16/2017",
#              "5/31/2017")
# 
# 
# 
# harvest <- harvest %>% filter((permit %in% rm_jay))

harv_na <- apply(X = harvest, MAR = 2, FUN = function(x){any(is.na(x))})
data.frame(missing_values = harv_na)


permit_na <- apply(X = permits, MAR = 2, FUN = function(x){any(is.na(x))})
data.frame(missing_values = permit_na)



```


## Tables of Counts


First we will look at some tables to get a sense of the data we are starting with.

Counts by mailing status:
```{r counts, echo=FALSE}

mail_cnts <- permits %>% group_by(mailing) %>% summarise(count = n())
knitr::kable(mail_cnts)


```

Counts by combinations of mailing and reporting status:
```{r}
status_cnts <- permits %>% group_by(mailing, status) %>% summarise(count = n())
knitr::kable(status_cnts)

```
```{r}

status_cnts <- status_cnts %>% filter(status != "BLANK REPORT")

status_cnts[1,3] <- 634
status_cnts[3,3] <- 238
```

## Important Quantities in Formulas

*Important: A "noncompliant" household is interpeted as a household with a mailing status of either 2 or 9,
and "compliant households" have a mailing status of 0 or 1. I also assumed each household had only one permit and one mailing status.

```{r}
# number of noncompliant (mailing == 2) households responding to the second reminder that reported fishing
ndf <- status_cnts %>% filter(mailing == 2, status == "HARVEST REPORTED") %>% ungroup() %>% select(count) %>% as.integer()


# total number of noncompliant households responding to the second reminder (mailing == 2)
nd <- status_cnts %>% filter(mailing == 2) %>% ungroup() %>% select(count) %>% sum() %>% as.integer()
Nd <- status_cnts %>% filter(mailing %in% c(2,9)) %>% ungroup() %>% select(count) %>% sum() %>% as.integer()

# estimated proportion of non-respondents that fished
w_hat <- ndf/nd


N <- sum(mail_cnts$count)


# Ncf2 <- status_cnts %>% filter((mailing == 0 | mailing == 1) & status == "HARVEST REPORTED") %>% ungroup() %>% select(count) %>% sum()
# Ncz2 <- status_cnts %>% filter((mailing == 0 | mailing == 1)  & status == "DID NOT FISH") %>% ungroup() %>% select(count) %>% sum()
# Ndf2 <- (N - (Ncf2 + Ncz2))*w_hat
# Ndf2 <- ndf + Ndf


# counts of all the people we know did or did not fish
# Ncf <- status_cnts %>% filter((mailing == 0 | mailing == 1 | mailing == 2) & status == "HARVEST REPORTED") %>% ungroup() %>% select(count) %>% sum()
# Ncz <- status_cnts %>% filter((mailing == 0 | mailing == 1 | mailing == 2)  & status == "DID NOT FISH") %>% ungroup() %>% select(count) %>% sum()



# N - Ncf - Ncz = 283, equals 279 non-respondents plus 4 blank reports
# Ndf is the estimated number of mailing = 9 that fished


Ndf <- Nd*w_hat
# Ndf <- (N - Ncf - Ncz)*w_hat
# 
# Ndz <- N - Ncf - Ncz - Ndf




ndf_d <- "number of households with mailing = 2 and status = 'harvest reported'"

nd_d <- "total number of households with mailing = 2"

w_hat_d <- "ndf/nd"

N_d <- "count of all households that recieved mailing status of either 0, 1, 2, or 9"

# Ncf_d <- "count of households that had status = 'harvest reported' and mailing status 0 or 1"

# Ncz_d <-  "count of households that had status = 'did not fish' and mailing status 0 or 1"

Ndf_d <- "(N - (Ncf + Ncz))*w_hat, estimated count of mailing = 9 households that fished"

# Ndz_d <- "N - Ncf - Ncz - Ndf, estimated count of mailing = 9 that did not fish"

# metric_tabl <- data.frame(Metric = c("ndf", "nd", "w_hat", "N", "Ncf", "Ncz", "Ndf", "Ndz"),
#                           Value = c(ndf, nd, w_hat, N, Ncf, Ncz, Ndf, Ndz),
#                           Description = c(ndf_d, nd_d, w_hat_d, N_d, Ncf_d, Ncz_d, Ndf_d, Ndz_d))
# metric_tabl$Value <- round(metric_tabl$Value, 2)
# 
# 
# knitr::kable(metric_tabl)

```


Next I joined the harvest data and the permit data by permit number. Shrimp and pot days become the totals (sum) for that permit. I did this so I'd have a dataframe with permit number, pot_days, shrimp gallons, mailing status and harvest status all together. The dataframe looks like this:
```{r}


permit_dat <- full_join(harvest, permits, by = "permit") %>% group_by(permit) %>%
  summarize(pot_days = sum(pot_days), shrimp = sum(shrimp), mailing = unique(mailing), status = unique(status), reported_fishing = unique(reported_fishing))

permit_dat
# permit_dat$pot_days[which(is.na(permit_dat$pot_days))] <- 0
# permit_dat$shrimp[which(is.na(permit_dat$shrimp))] <- 0
# PROBLEM! 
# unique(permit_dat$status)



```

This results in a dataframe with one row per permit in the permit records data sheet. There were `r length(unique(permits$permit))` permits in the permit excel sheet, and `r length(unique(harvest$permit))` in the harvest sheet.

More quantities used in the final formulas:

```{r}
# calculate harvest and effort for compliant households.
Hcf_gals <- permit_dat %>% filter((mailing == 0 | mailing == 1)  & !is.na(shrimp)) %>% select(shrimp) %>% sum()
Hcf_days <- permit_dat %>% filter((mailing == 0 | mailing == 1)  & !is.na(pot_days)) %>% select(pot_days) %>% sum()


# calculate mean harvest and effort from those that responded to the second request
hdf_gals<- permit_dat %>% filter(mailing == 2, !is.na(shrimp), status == "HARVEST REPORTED") %>% select(shrimp) %>% summarize(mean = mean(shrimp)) %>% as.numeric()
hdf_days <- permit_dat %>% filter(mailing == 2, !is.na(pot_days), status == "HARVEST REPORTED") %>% select(pot_days) %>% summarize(mean = mean(pot_days)) %>% as.numeric()


# Should the variance formulas in equations 12 and 13 in the op plan have hats on them?
VAR_gals <- permit_dat %>% filter(mailing == 2, !is.na(shrimp), status == "HARVEST REPORTED") %>% select(shrimp) %>% summarize(sd = var(shrimp)) %>% as.numeric()
VAR_days <-  permit_dat %>% filter(mailing == 2, !is.na(pot_days), status == "HARVEST REPORTED") %>% select(pot_days) %>% summarize(sd = var(pot_days)) %>% as.numeric()


# multiply the means of respondents of the second request by the number that failed to respond
# to get an estimate of the total harvest and effort by non-respondents

Hdf_gals <- (Ndf)*hdf_gals
Hdf_days <- (Ndf)*hdf_days



Name <- c("Hcf", "hdf", "VAR", "Hdf")
Desc <- c("Total for households with mailing = 0, 1, or 2",
          "Mean for mailing = 2 AND status = 'HARVEST REPORTED'",
          "Variance for mailing = 2 AND status = 'HARVEST REPORTED'",
          "Ndf x hdf")

Value_gal <- round(c(Hcf_gals, hdf_gals, VAR_gals, Hdf_gals))
Value_days <- round(c(Hcf_days, hdf_days, VAR_days, Hdf_days))

estimates_df <- data.frame(Estimate = Name, Description = Desc, Value_gal, Value_days)


knitr::kable(estimates_df)
```


## Final Estimates

```{r}

tot_gal <- Hcf_gals + Hdf_gals
tot_days <-  Hcf_days + Hdf_days


# I think ndf/Ndf is incorrect because its larger than 1 

var_hdf_gal <- (Ndf-ndf)/(Ndf-1)*(VAR_gals/ndf)
var_hdf_days <- (Ndf-ndf)/(Ndf-1)*(VAR_days/ndf)

var_what <- (Nd - nd)/(Nd-1)*(w_hat*(1-w_hat)/(nd-1))
var_Ndf <- Nd^2*var_what

total_var_gal <- Ndf^2*var_hdf_gal + hdf_gals^2*var_Ndf - var_hdf_gal*var_Ndf
total_var_days <- Ndf^2*var_hdf_days + hdf_gals^2*var_Ndf - var_hdf_days*var_Ndf


se_gal <- round(sqrt(total_var_gal), 2)
se_days <- round(sqrt(total_var_days), 2)


knitr::kable(data.frame(`Estimated Totals` = c(tot_gal, tot_days), `Standard Errors` = c(se_gal, se_days)))




```

