---
title: "Tanner Expansion 2019"
author: "Ben Buzzee"
date: "August 23, 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)


harvest <- read.csv("harvest_records_19.csv", header = T, stringsAsFactors = F)
permits <- read_csv("permit_records_19.csv", n_max = 1681)



# change all variable names to lower (personal preference)
names(harvest) <- tolower(names(harvest))
names(permits) <- tolower(names(permits))



harvest <- harvest %>% select(harvestpersonalinfoid, ontimereport, reportdate, permitno, permittypeid, harvestcount, didyoufish, mailing, harvestcount, areacode)
permits <- permits %>% select(permittypeid, `reported?`, name_combi)


names(permits) <- c("type", "reported", "name")

# replace with yes/no
permits$reported <- gsub(".*no.*", "no", permits$reported)
permits$reported <- gsub(".*yes.*", "yes", permits$reported)


# determine whether mailed in reports were late (old mailing = 2)
harvest$reportdate <- as.Date(harvest$reportdate, format = "%m/%d/%y")
harvest$late <- ifelse(harvest$reportdate < as.Date("2019-3-1"), "yes", "no")
harvest$areacode <- ifelse(harvest$areacode == "", NA, ifelse(harvest$areacode %in% c("A", "B", "C"), "ABC", harvest$areacode))

```


Total reported harvest by permit type and area:
```{r}
total <- harvest %>% group_by(areacode) %>% summarize(crab_count = sum(harvestcount, na.rm = T))


knitr::kable(total)

```






Total number of non-respondents:

```{r}
permits %>% group_by(type, reported) %>% summarize(count = n())
```


So we see that `r 139 + 6 + 21` people failed to respond. 


```{r}

# non-responding households that reported fishing
non_rspd_sport <- 139 + 6

# total nonresponding households
non_rspd_sub <- 21 + 6 

N <-non_rspd_sport+ non_rspd_sub
```


Next, we will estimate the number of non-respondents that fished:

First we need to find the proportion of late responders that fished in each area:
```{r}

harvest <- harvest %>% filter(didyoufish == TRUE)

late_harv <- harvest %>% group_by(areacode, late) %>% summarize(count= n())

late_harv
```


So we will assume the proportion of non-responders that fished in each area is the same as the proportion of late responders that fished in each area.

```{r}

# late responders in each area that did report fishing
ndf_E <- 654

ndf_D <- 67

ndf_ABC <- 81

ndf <- c(ndf_ABC, ndf_D, ndf_E)
# total late responders
nd <- 654 + 67 + 81


w_hat_E <- ndf_E/nd
w_hat_D <- ndf_D/nd
w_hat_ABC <- ndf_ABC/nd
prop_non_rspd_fished = c(w_hat_ABC,w_hat_D, w_hat_E)

proportions <- data.frame(area = c("ABC", "D", "E"), prop_non_rspd_fished)


knitr::kable(proportions)

```



Next, we'll find the mean harvest per area:

```{r}
mean_catches <- harvest %>% filter(late == "yes") %>% group_by(areacode) %>% summarize(mean = mean(harvestcount, na.rm = T))
mean_catches


```

Finally, to find our expanded harvest numbers, we will multiply total non-respondents, by the porportion we expected to fish in each area, times the average catch



```{r}
knitr::kable(cbind(areacode = c("ABC", "D", "E"), crab_count = round(172*proportions$prop_non_rspd_fished*mean_catches$mean, 0)))
```

Next we'll estimate standard errors for the above estimates:

```{r}
var_xhat <- harvest %>% filter(late == "yes") %>% group_by(areacode) %>% summarize(var(harvestcount, na.rm = T))



var_phat <- prop_non_rspd_fished*(1 - prop_non_rspd_fished)/nd

N <- 172


SE <- sqrt(N^2*(prop_non_rspd_fished^2*var_xhat$`var(harvestcount, na.rm = T)` + mean_catches$mean^2*var_phat))

se_df <- data.frame(cbind(areacode = c("ABC", "D", "E"), stderr = round(SE, 0)))

knitr::kable(se_df)
```

